---
title: "Chica"
output: html_notebook
---

## 05/05/20 ----------------------------------------------------------------------
## 05/08/20 ----------------------------------------------------------------------

By Friday: data in order, choose what algorithms to work on
Next week: Work on our algorithms

Options:

* lm [ lm(covid cases ~ ., data = covid.eviction) ]
    supervised
    predictive
    linear
    numeric outcome

* k-means clustering
    unsupervised
    we can visualize clusters of similar data by distance

* knn
    supervised
    predictive
    can predict based on nearest data points (categorial predictor)
    
* decision trees
    supervised & unsupervised
    predictive:
      regression context: predict covid cases using features like eviction
      can use lm or other models as predictors
    classification:
      can just be used to visualize by similar features
    
* random forest
    supervised & unsupervised
    better tree models
    predictive
    classification
    
* lda
    unsupervised
    used for clustering: maxes distance between categories, mins distance within
    mostly just for grouping similar things

* pca
    unsupervised
    transforms data to new coordinate system which maxes variance between classes
    also mostly used just to make data easy to visualize by similarity/difference
    kind of difficult with our data, but we could look at just eviction similarities/differences and how that translates to covid

* splines
    fits a bunch of data with a squiggly line
    requires a more or less clear pattern, right? and continuous data on both axes?

* hierarchical clustering
    

* kernel smoothing
    supervised
    predictive
    estimates function as weighted avg of neighboring data (closer points weighted more)



## Unsupervised:


## Supervised:


## 05/07/20 ----------------------------------------------------------------------

Libraries:
```{r libraries, include=FALSE}
library(readxl)
library(tidyverse)
```

Load in data:
(covid data for 4/30/20)
```{r data, eval=FALSE, include=FALSE}
covid.data <- read.csv("covid_confirmed_usafacts_2.csv")
covid.cases <- covid.data[, c(1:4, 104)]
colnames(covid.cases)[c(1, 2, 5)] <- c("countyFIPS", "County", "covidCases")
```

just in case it's useful I'm gonna bring in death numbers.
```{r, eval=FALSE, include=FALSE}
covid.deaths.data <- read_csv('covid_deaths_usafacts.csv')
covid.deaths <- covid.deaths.data[, c(1:4, 104)]
colnames(covid.deaths)[c(1, 2, 5)] <- c("countyFIPS", "County", "covidDeaths")
```

Pop data we can use to control/adjust covid case and eviction filing rate data.
```{r, eval=FALSE, include=FALSE}
covid.pop.data <- read_csv('covid_county_population_usafacts.csv')
covid.pop <- covid.pop.data
```

combine all of the covid data
```{r, eval=FALSE, include=FALSE}
covid <- full_join(covid.cases, covid.deaths[, c(1, 5)],
                   by = "countyFIPS")

covid <- full_join(covid, covid.pop[, c(1, 4)],
                   by = "countyFIPS")

```

filtering out non-county rows ("Statewide Unallocated")
```{r, eval=FALSE, include=FALSE}
covid <- covid %>%
  filter(countyFIPS != 0)
covid <- covid %>%
  filter(countyFIPS != 1)
```

this is at a bunch of levels but not county I think so I'm going to get something at the county level. that way it can match our county-level covid data.
```{r, eval=FALSE, include=FALSE}
# eviction.data <- read.csv("eviction.csv")
# eviction <- eviction.data
```

this is just the state-reported filing rate for several years. I'll narrow it down to the most recent year.
```{r, include=FALSE}
# eviction.county.data <- read_xlsx('ExtStatsFull.xlsx', sheet = 4)
# eviction <- eviction.county.data
```

Let's see what the most recent year looks like:
```{r, eval=FALSE, include=FALSE}
eviction.2016 <- eviction %>%
  filter(year == 2016)
```

Trying to figure out why there are more rows in covid than in eviction data. I guess we can just go with counties where we have both info for both. This leaves us with 26 states.
```{r, eval=FALSE, include=FALSE}
length(unique(covid$countyFIPS))
length(unique(eviction.2016$fips))
  
covid.eviction <- inner_join(eviction.2016[, c(1, 4, 5)], covid,
          by = c("fips" = "countyFIPS"))
colnames(covid.eviction)[3] <- "Evictions"
```

Create our mean benchmarks for evictions, deaths, and cases, all over pop. 
US total pop taken from https://www.census.gov/popclock/
```{r, eval=FALSE, include=FALSE}
state.mean.evictions <- covid.eviction %>%
  group_by(State) %>%
  summarize(state_mean_filing_rate = mean(as.numeric(`eviction-filing-rate`), na.rm = TRUE))

covid.eviction <- full_join(covid.eviction, state.mean.evictions,
                            by = "State")

covid.eviction  <- covid.eviction %>%
  mutate(evictions_greater_state = ifelse(`eviction-filing-rate` > state_mean_filing_rate,
                                          1,
                                          0))

US.mean.evictions <- covid.eviction %>%
  summarize(US_mean_filing_rate = mean(as.numeric(`eviction-filing-rate`), na.rm = TRUE))

covid.eviction <- covid.eviction %>%
  mutate(US_mean_filing_rate = US.mean.evictions$US_mean_filing_rate)

covid.eviction <- covid.eviction %>%
  mutate(evictions_greater_US = ifelse(`eviction-filing-rate` > US_mean_filing_rate,
                                       1,
                                       0))
```

I'm gonna base covid data on average of all states/counties including those that we don't have evictiond data for.
```{r, eval=FALSE, include=FALSE}
Us.covid.cases.rate <- covid.eviction %>%
  summarize(US_covid_cases_rate = sum(covidCases)/329584132)

covid.eviction <- covid.eviction %>%
  mutate(US_covid_cases_rate = Us.covid.cases.rate$US_covid_cases_rate)

US.covid.death.rate <- covid %>%
  summarize(US_covid_death_rate = sum(covidDeaths)/sum(covidCases))

covid.eviction <- covid.eviction %>%
  mutate(US_covid_death_rate = US.covid.death.rate$US_covid_death_rate)

covid.eviction <- covid.eviction %>%
  mutate(county_covid_cases_rate = covidCases/population)

covid.eviction <- covid.eviction %>%
  mutate(county_covid_death_rate = covidDeaths/covidCases)

State.covid.cases.rate <- covid.eviction %>%
  group_by(State) %>%
  summarize(state_covid_cases_rate = mean(county_covid_cases_rate))

covid.eviction <- full_join(covid.eviction, State.covid.cases.rate,
                            by = "State")

State.covid.death.rate <- covid.eviction %>%
  group_by(State) %>%
  summarize(state_covid_death_rate = mean(county_covid_death_rate))

covid.eviction <- full_join(covid.eviction, State.covid.death.rate,
                            by = "State")
  
covid.eviction <- covid.eviction %>%
  mutate(covidCases_greater_state = ifelse(county_covid_cases_rate > state_covid_cases_rate,
                                           1,
                                           0))

covid.eviction <- covid.eviction %>%
  mutate(covidCases_greater_US = ifelse(county_covid_cases_rate >
                                          US_covid_cases_rate,
                                        1,
                                        0))

covid.eviction <- covid.eviction %>%
  mutate(covidDeaths_greater_US = ifelse(county_covid_death_rate >
                                          US_covid_death_rate,
                                        1,
                                        0))

covid.eviction <- covid.eviction %>%
  mutate(covidDeaths_greater_state = ifelse(county_covid_death_rate >
                                              state_covid_death_rate,
                                            1,
                                            0))
```

Reorganize columns:
```{r, eval=FALSE, include=FALSE}
covid.eviction <- covid.eviction[, c(1:2, 4, 9, 6, 5, 3, 10:11, 14, 12:13, 
                                     7:8, 17:20, 15:16, 21, 24, 22:23)]
```

write new csv:
```{r, eval=FALSE, include=FALSE}
# write_csv(covid.eviction, "covid_eviction.csv")
```

## 05/08/2020 --------------------------------------------------------------------

Add in other evictionlab variables:
```{r}
covid.eviction <- read_csv('covid_eviction.csv')

US.evictions <- read_csv('counties_us.csv')

eviction <- US.evictions

eviction.2016 <- eviction %>%
  filter(year == 2016)

length(unique(covid$countyFIPS))
length(unique(eviction.2016$GEOID))

str(eviction.2016$GEOID)
str(covid$countyFIPS)

eviction.2016$GEOID <- as.numeric(eviction.2016$GEOID)
  
covid.eviction <- inner_join(eviction.2016[, -c(3:5)], covid,
          by = c("GEOID" = "countyFIPS"))

# write_csv(covid.eviction, "covid_eviction2.csv")

```



